10 years of school closures and consolidations: the impact on school accessibility

### 2025-01-12 paper/Manuscript.tex
Proofread revisions

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\raggedbottom
\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\abstract{Reducing motorized travel and increasing active travel are top
priorities for many communities. However, these priorities are
confronted with pressures to reduce the fiscal burden of public
services, including schools. Public school boards have been strained to
do more with less, and opting to close under-capacity schools to deliver
educational services more cost-effectively. However, once a school is
closed, students' proximity to schools decreases, and those who
previously lived close enough to walk or cycle can no longer do so. In
this paper, the City of Hamilton, a mid-sized city in Ontario, Canada,
which experienced a 10\% decrease in elementary school locations between
2011 and 2021, serves as a case study. We quanitfiy how these closures
and consolidations have reshaped accessibility for the elementary
school-aged population and discuss their implications for active and
motorized travel. To measure accessibility, spatial availability, a
singly-constrained multimodal measure of spatial accessibility is used.
This method's proportional allocation mechanism enables interpretable
comparison between accessibility `before' and `after' the implementation
of school closure/consolidation policy considering changes in potential
travel time and population. The analysis is conducted at the parcel
level, considering all school seats as eligible and incorporating
observed home-to-school travel patterns for motorized and non-motorized
modes. We present system-wide impact estimates resulting from these
closures. The findings reveal that overall spatial availability in the
city declined. While the school consolidation policy may have achieved
cost savings in facility operation and maintenance, the burden falls on
students, families, and communities through \emph{potentially} reduced
opportunities for physical activity through an active journey to school
and increased motorized travel externalities.}

@@

The amount of services accessible within communities oscillate: economic
and demographic shifts, rural and urban configurations and political
in/action all induce change. These forces spur pressures to
reorganization and consolidation of public facilities that provide
services, potentially improving access for some but not for all
(Christiaanse 2020; Rosik, Puławska-Obiedowska, and Goliszek 2021).
Schools are a compelling case study in this respect. In many communities
globally, school closure and consolidation policies have been pursued by
governments to optimize per-student operational cost-efficiency (Rong et
al. 2022; Dai et al. 2019). Research has highlighted that these polices
have knock-on implications, including the weakening of communities'
social infrastructure (Butler, Kane, and Cooligan 2019; Irwin and
Seasons 2012; Autti and Hyry-Beihammer 2014) and negative
household-level fiscal impacts such as a reduction of property values
(Merrall, Higgins, and Páez 2024). Operationally, savings may be
experienced but they are paid at different levels by students, families,
communities and society.

A school offers many things to a community: but one dimension is the
travel to and from school, often overlooked by decision-makers
considering school closure and consolidation policies (Bierbaum, Karner,
and Barajas 2021; J. Lee and Lubienski 2017). Schools and residential
locations often proximately co-located, the journey-to-school is an
excellent opportunity for physical activity (Desjardins et al. 2024).
However, closures exchange the physical activity of students' realized
active trips or ``reserved'' active trips (Morency, Roorda, and Demers
2009; Morency, Demers, and Lapierre 2007) (e.g., trips that could be
made active) for motorized travel (Rong et al. 2022). Indeed, motorized
travel to school is the largest contributor of carbon emissions from
personal daily travel after the journey-to-work (Rong et al. 2022;
Pantelaki, Claudia Caspani, and Maggi 2024). Despite these societal
implications, many communities are closing and/or consolidating schools
(Sageman 2022). For instance, 65\% of comprehensive schools (e.g.~grades
1 through 9) in rural Finland since 1990 closed (Autti and
Hyry-Beihammer 2014), 10\% of elementary schools (kindergarten through
grade 8) in Chicago, USA in 2013 (J. Lee and Lubienski 2017), and 7\% of
primary and lower secondary schools (kindergarten through grade 9) were
closed in Denmark between 2010-2011 (Beuchert et al. 2018).

In this paper, we explore the case study of Hamilton, a mid-sized city
in Canada that has pursued similar school closure and consolidation
policies. We detail the Canadian political context that incentived these
policies and quantify their transportation implications using data that
constructs where students may have lived and the capacity of schools in
2011, 2016 and 2021. To quantify transportation implications, we use a
novel multimodal extension of a singly-constrained accessibility measure
known as spatial availability (Soukhov et al. 2023; Soukhov et al.
2024). Spatial availability is used to estimate active (i.e., walking)
and motorized (i.e., car) accessibility considering the capacity of each
school, and the residential location and quantity of school-aged
population, and travel times from residential locations to schools for
three time periods. In this way, this work contributes a empirical
findings documenting how government austerity hurts public education,
along with a methodology that results in interpretable accessibility
values that can be compared across modes, the City, and three time
periods.

To foreshadow the paper's conclusions, our findings indicate that school
seats have became less spatially available in the city. This change
happened unevenly, with certain communities being disproportionately
affected. The school closure and consolidation policy pursued may have
achieved cost savings in facility operation and maintenance. However, it
has come at a cost to families, students, and communities, resulting in
lost opportunities for healthier active travel and a potential increase
in motorized travel. We investigate: were the closures and
consolidations worth the operational savings? The methods used in this
paper can be applied by decision-makers to evaluate the spatial and
transport-related impacts of public service provision policy.
Furthermore, to increase the diffusion of this analysis, all work is
transparent and reproducible, following best practices in the spatial
sciences (Brunsdon and Comber 2021; Antonio Páez 2021). Accordingly, all
associated code and data is publicly available within the lead author's
\href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
repository}.

@@

The examination of access to public services is a well established area
of the transportation literature: from parks (El-Murr et al. 2021) to
healthcare (Pereira, Braga, et al. 2021) and associated dimensions of
inequities (Tiznado-Aitken, Munoz, and Hurtubia 2021; Bittencourt and
Giannotti 2023; Kelobonye et al. 2020), spatial accessibility has been
explored in cases studies worldwide.

Accessibility to schools has been a public service explored across
several transportation dimensions. For instance, Marques, Wolf, and
Feitosa (2020) demonstrates that lower-income neighbourhoods, compounded
by other intersecting dimensions of disadvantage, tend to have lower
accessibility to schools compared to more advantaged neighbourhoods in
Portugal. Similar conclusions are drawn in West Virginia, US by Talen
(2001), England by Burgess et al. (2011), Baton Rouge, US by Williams
and Wang (2014), São Paulo and Curitiba, Brazil by Moreno-Monroy,
Lovelace, and Ramos (2018), Pizzol, Giannotti, and Tomasiello (2021),
and Bittencourt and Giannotti (2023), and Santiago, Chile by
Tiznado-Aitken, Munoz, and Hurtubia (2021). Research has also
demonstrated that when school closures occurs, what schools end up
closing are not a politically neutral decision. For instance, these
spatial trends are explored in the context of England by Pinch (1987),
areas of declining school-aged populations in Dreseden, Germany by
Müller (2011), rural Finnish communities by Autti and Hyry-Beihammer
(2014), and Chicago, US by J. Lee and Lubienski (2017).

The methods applied to measure school accessibility have been varied,
from simple indicators to complex metrics. Beginning from more simple,
Marques, Wolf, and Feitosa (2020) present indicators of school seats per
student along with the shortest distance from origin to schools within a
school catchment, methods also used in Talen (2001) and Burgess et al.
(2011). Authors make refinements to these methods; Müller (2011)
considers travel time by public transit from origins to schools and
Pizzol, Giannotti, and Tomasiello (2021) considers travel times by
different modes and multiple schools from a single origin along with
other indicators related to school quality. By contrast, Williams and
Wang (2014), J. Lee and Lubienski (2017), Moreno-Monroy, Lovelace, and
Ramos (2018), Tiznado-Aitken, Munoz, and Hurtubia (2021), and
Bittencourt and Giannotti (2023) use more complex approaches such as the
two-step floating catchment area (2SFCA). The 2SFCA is popular in
measure in evaluating access to healthcare services (Antonio Páez,
Higgins, and Vivona 2019) and yields a supply-to-demand ratio that
accounts for the population demanding a service and the quantity that is
supplied within a certain distance range (Shen 1998; Luo and Wang 2003).
The consideration of the supply of service is pertinent in the case of
schools and is acknowledged to as a limitation to using simpler
accessibility measures (Pizzol, Giannotti, and Tomasiello 2021).

In addition to competition for school-seats, theoretically, the
advantage students using faster modes (e.g., motorized modes) offer in
terms of a wider range in school-seat choice is also important. Though
some of the reviewed works using 2SFCA approaches consider multiple
modes, explicit attention to the modal \emph{advantage} has not been
discussed. To this end, in this work we use spatial availability, a
singly-constrained measure of spatial accessibility (Soukhov et al.
2023; Soukhov et al. 2024). Spatial availability allows for the
representation of how many `school-seats' are spatially available to
motorized- and active-trip taking elementary school-aged student
population at a spatial-level, facilitating interpretable comparison
across years. Further, spatial availability can be represented as
school-seats per capita, mathematically equivalent to the
supply-to-demand ratio of the 2SFCA (see appendix in Soukhov et al.
(2023)).

@@

Hamilton is a mid-size city (\textasciitilde570,000 pop) with rural,
suburban and urban characteristics in the province of Ontario, Canada
(Government of Canada 2022a). The city's public English education is
covered by the Hamilton-Wentworth District School Board (HWDSB). In
2013, the HWDSB released its Long Term Facilities Master Plan (HWDSB
2013). To ensure ``equitable, affordable and sustainable learning
facilities'', the Master Plan indicated that 80\% of its elementary
schools (attended by students aged approximately 5 to 14) would be
subject to evaluations, including a ``Pupil Accommodation Review''
(HWDSB 2013). This Master Plan and the slated Pupil Accommodation
Reviews prompted public outcry, as the outcome of Pupil Accommodation
Reviews are typically schools closures and consolidations (Craggs 2013;
Seasons 2014b).

Like all school boards in Ontario, HWDSB relies on funding from the
province which is distributed through a formula based on pupil enrolment
(Mackenzie 2018; Irwin and Seasons 2012). However, it has been
demonstrated that this funding formula has become insufficient to
maintain schools in a state of good repair (Auditor General of Ontario
2015). Further, existing ``top-up'' programs to assist in operation and
maintenance costs (e.g., School Facilities Operation and Renewal Grant)
were entirely phased out by 2018 (TDSB 2024). Within this funding
backdrop, the Pupil Accommodation Review guidelines outlined a
``community-focused'' process to assessing the value that a school
provides to students, the community, and the local economy (Seasons
2014b; Ministry of Education 2006). However in practice, critical
residents felt the outcomes of the Pupil Accommodation Reviews were
predetermined (Thompson, Collins, and Dean 2024): older schools were
slated to be closed, and a wholesome breakdown of costs to repair and
maintain schools were not made available (Kleinhuis 2013). Other critics
pointed to unbalanced incentives motivating the HWDSB: ``By closing
three schools, the HWDSB would have a strong case with the Ministry of
Education to receive full funding for a new school'' (Kleinhuis 2013).
The Ministry's framework promotes the closing and consolidation of
schools rather than cost-effective solutions for keeping older schools
open (Kleinhuis 2013).

The HWDSB and the Hamilton-Wentworth Catholic District School Board
(HWCDSB), which together provide public English education to the
majority of school-aged children in Hamilton (FAO 2023), were
particularly impacted. While Accommodation Reviews were a province-wide
trend during this period, Hamilton's school boards underwent an
unprecedented number of reviews compared to other boards in other
Ontario municipalities. (Seasons 2014a). In the wave of Pupil
Accommodation Reviews between 2013 and 2016, 12 of the 147 elementary
schools closed in Hamilton. Likely in part due to the public
unpopularity of school closures, in 2017, the provincial government
introduced a moratorium on beginning new Pupil Accommodation Reviews. In
the following years, the HWDSB addressed its backlog and closed 12 more
schools between 2017 to 2021 (HWDSB 2023). Though elementary school
locations declined by 10\% between 2011-2021, HWDSB intends to conduct
future Pupil Accommodation Reviews as initially scheduled, waiting until
the end of the moratorium before dates are assigned (HWDSB 2023).

A justification in these Pupil Accommodation Reviews are operational
savings based on ``projected'' reductions in student population and
``under-utilized'' school capacity (Craggs 2012; TDSB 2024; OPSBA 2024).
As later demonstrated in this study, these policies in fact did result
in school capacity per student reducing overall. Indeed, school seats in
the city declined 5\%. It should be noted that though closures dominated
the Pupil Accommodation Review process, some schools were
``consolidated'' e.g., a school or two were closed and another school
was expanded or opened to accommodate students from closed schools.
Taken that school locations decreased more than overall school-seats,
elementary schools in Hamilton are on average larger than they once
were.

In 2024, the moratorium on Pupil Accommodation Reviews is still in place
and school boards are motivated by the same funding conditions. In
absence of new revenue-generating levers and under the same funding
formula with even lower inflation adjusted per-pupil funding, many
school boards are calling for the end of the moratorium (OPSBA 2024;
TDSB 2024; Draaisma 2024). In the meantime, some school boards are
reducing specialized and community-wide programming (TDSB 2024). Under
the status-quo, the end of the moratorium will result in further school
closures. These pertinent decisions, made under an austere funding
formula and flawed school board incentives does not create an
environment where solutions for a sustainable public education system
can thrive.

Though schools have a plethora of benefits not captured by their
maintenance and operational costs (Butler, Kane, and Cooligan 2019;
Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014), this paper's
scope is limited to their travel implications. For students who live in
proximity to schools, their closure or consolidation directly eliminate
the realized or potential active trips that are, or could have been,
made. Closing a school permanently reduces the potential for active
trips and further entrenches reliance on motorized travel. Additionally,
from the perspective of modal equity, school-aged populations with
access to motorized transport gain a speed-time advantage, providing
them with greater flexibility and choice in accessing school-seats.

The spatial and population-level distribution of these impacts are of
question for this paper. Specifically, the paper's aim is to demonstrate
how these policies impacted the motorized and non-motorized (active)
school-seat accessibility landscape. We hypothesize that the school
closure and consolidation policies reduced accessibility to school-seats
for many students. By extension, this loss in access and the
displacement of active trips likely resulted in decreased potential
active travel minutes and increased motorized travel minutes, with these
impacts distributed inequitably across space and populations. To test
these hypotheses, the paper examines the following:

\begin{itemize}
\tightlist
\item
  How has spatial availability changed for home-to-school trips
  following school closures and consolidations?
\item
  What is the extent of the loss in walking minutes for these trips?
\item
  How has motorized travel increased as a result?
\item
  Which communities, particularly households with a high prevalence of
  low-income and hence more likely experiencing transport poverty,
  experienced the greatest losses due to this policy?
\end{itemize}

@@

First, elementary schools locations, how the changed throughout the
study period and their capacities are required for each studied year.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\raggedbottom
\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\abstract{Reducing motorized travel and increasing active travel are top
priorities for many communities. However, these priorities clash against
many government's reluctance to invest in public services, including
schools. Public school boards are increasingly asked to do more with
less, a directive often accomplished by closing and consolidating
schools. Alas, this policy tends to reduce the proximity of schools for
many students, thus reducing the chances for active travel. In this
paper we study the City of Hamilton, a mid-sized city in Ontario,
Canada, which experienced numerous elementary schools closures between
2011 and 2021. Our analysis evaluates how school closures and
consolidations reshaped accessibility for elementary school students by
motorized and active travel. In this task we use spatial availability, a
singly-constrained multimodal measure of spatial accessibility. This
method's proportional allocation mechanisms allow us to make consistent
comparisons in accessibility between time periods. Analysis is conducted
at the parcel level, considering on-the-ground-capacity (OTGC) of
schools and observed home-to-school trip lengths for motorized and
active modes. Our findings reveal that overall spatial availability in
the city declined, and while the school consolidation policy pursued by
the province may have achieved cost savings in facility operation and
maintenance, it imposes ongoing costs on students, families, and
communities through reduced opportunities for physical activity and
entrenched reliance on motorized travel.}

@@

Public services accessible within communities tend to oscillate:
economic and demographic shifts, rural and urban configurations and
political in/action all induce change. These forces spur pressures to
reorganize and consolidate public facilities, potentially improving
access for some but not for all residents in a community (Christiaanse
2020; Rosik, Puławska-Obiedowska, and Goliszek 2021). Public schools are
a compelling case study in this respect. Across the globe, school
closure and consolidation policies have been pursued by governments to
reduce the cost of running public education systems (Rong et al. 2022;
Dai et al. 2019). Research has highlighted that these polices have
knock-on implications, including the weakening of communities' social
infrastructure (Butler, Kane, and Cooligan 2019; Irwin and Seasons 2012;
Autti and Hyry-Beihammer 2014) and negative household-level fiscal
impacts such as a reduction of property values (Merrall, Higgins, and
Páez 2024). Operationally, savings may be experienced by governments,
but they are paid at different levels by students, families,
communities, and ultimately society in general.

A school offers many things to a community, and one relevant dimension
is how it shapes travel. This dimension is often overlooked by
decision-makers tasked with implementing school closure and
consolidation policies (Bierbaum, Karner, and Barajas 2021; J. Lee and
Lubienski 2017). When schools and residential locations are co-located,
the journey-to-school presents an excellent opportunity for physical
activity (Desjardins et al. 2024). However, school closures exchange the
opportunities for active trips by students (Morency, Roorda, and Demers
2009; Morency, Demers, and Lapierre 2007) for the near certainty of
motorized travel (Rong et al. 2022). Indeed, motorized travel to school
is the largest contributor of carbon emissions from personal daily
travel after the journey-to-work (Rong et al. 2022; Pantelaki, Claudia
Caspani, and Maggi 2024). Alas, these impacts are not often considered
in policy when schools close and/or consolidate (Sageman 2022), a
relatively common occurrence globally. For instance, 65\% of
comprehensive schools (e.g.~grades 1 through 9) in rural Finland since
1990 closed (Autti and Hyry-Beihammer 2014), 10\% of elementary schools
(kindergarten through grade 8) in Chicago, USA in 2013 (J. Lee and
Lubienski 2017), and 7\% of primary and lower secondary schools
(kindergarten through grade 9) were closed in Denmark between 2010-2011
(Beuchert et al. 2018).

In this paper, we investigate the case of Hamilton, a mid-sized city in
Canada that has been forced by the province of Ontario to pursue school
closure and consolidation policies. We quantify the transportation
implications of these policies using the capacity of schools in 2011,
2016 and 2021, after several waves of school closures and
consolidations. To this end, we use a novel multimodal extension of a
singly-constrained accessibility measure known as spatial availability
(Soukhov et al. 2023; Soukhov et al. 2024). We use this measure to
estimate active (i.e., walking) and motorized (i.e., car) accessibility
while constraining the supply of the service to the capacity schools. In
this way, this work furnishes evidence to document how government
austerity hurts public goals (such as physical activity) and locks
school districts and families into motorized patterns of travel for the
foreseeable future.

To foreshadow the paper's conclusions, our findings indicate that school
seats have became less available in the city. This change happened
unevenly, with certain communities being disproportionately affected.
The school closure and consolidation policy pursued by the government
may have achieved cost savings in facility operation and maintenance.
However, it has come at a cost to families, students, and communities,
resulting in lost opportunities for healthier life-styles and an
inevitable increase in motorized travel (Hammer and Baluja 2012; Rose
2013; BUCK 2019). To increase the diffusion of this analysis, all work
is transparent and reproducible, following best practices in the spatial
sciences (Brunsdon and Comber 2021; Antonio Páez 2021). For this reason,
all associated code and data are publicly available within the lead
author's
\href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
repository}.

@@

The examination of accessibility to public services is well-established,
including studies of parks (Reyes, Paez, and Morency 2014; El-Murr et
al. 2021), healthcare (Pereira, Braga, et al. 2021), and public
transportation (Moniruzzaman and Paez 2012; Tiznado-Aitken, Munoz, and
Hurtubia 2021), among many other services (e.g., Bittencourt and
Giannotti 2023; Kelobonye et al. 2020).

Schools are a service that has been studied across several
transportation dimensions. For instance, Marques, Wolf, and Feitosa
(2020) demonstrate that lower-income neighbourhoods in Portugal tend to
have lower accessibility to schools compared to more advantaged
neighbourhoods. Similar conclusions are drawn in West Virginia, US by
Talen (2001), England by Burgess et al. (2011), Baton Rouge, US by
Williams and Wang (2014), São Paulo and Curitiba, Brazil by
Moreno-Monroy, Lovelace, and Ramos (2018), Pizzol, Giannotti, and
Tomasiello (2021), and Bittencourt and Giannotti (2023), and Santiago,
Chile by Tiznado-Aitken, Munoz, and Hurtubia (2021). Research has also
demonstrated that school closures are seldom politically neutral
decisions. For instance, spatial trends were explored in the context of
England by Pinch (1987), areas of declining school-aged populations in
Dreseden, Germany by Müller (2011), rural Finnish communities by Autti
and Hyry-Beihammer (2014), and Chicago, US by J. Lee and Lubienski
(2017).

The methods applied to measure school accessibility have also been
varied. For example, Marques, Wolf, and Feitosa (2020) present
indicators of school seats per student along with the shortest distance
from origin to schools within a school catchment. Similar methods are
also used in Talen (2001) and Burgess et al. (2011). Several authors
propose refinements to these methods; Müller (2011), for instance,
considers travel time by public transit from origins to schools, whereas
Pizzol, Giannotti, and Tomasiello (2021) considers travel times by
different modes and multiple schools from a single origin along with
other indicators related to school quality. By contrast, Williams and
Wang (2014), J. Lee and Lubienski (2017), Moreno-Monroy, Lovelace, and
Ramos (2018), Tiznado-Aitken, Munoz, and Hurtubia (2021), and
Bittencourt and Giannotti (2023) use more complex approaches such as the
two-step floating catchment area (2SFCA). The 2SFCA is popular in
measure in evaluating access to healthcare services (Antonio Páez,
Higgins, and Vivona 2019) and yields a supply-to-demand ratio that
accounts for the population demanding a service and the quantity that is
supplied within a certain distance range (Shen 1998; Luo and Wang 2003).
The consideration of the supply of service is pertinent in the case of
schools and is acknowledged to as a limitation to using simpler
accessibility measures (Pizzol, Giannotti, and Tomasiello 2021).

In addition to competition for school-seats, theoretically, the use of
faster modes (e.g., motorized modes) offers an advantage in terms of a
wider range in school-seat choice. Though some of the works reviewed
consider multiple modes, explicit attention to the modal
\emph{advantage} has not been discussed. To this end, in this work we
use spatial availability, a singly-constrained measure of spatial
accessibility (Soukhov et al. 2023; Soukhov et al. 2024). Spatial
availability allows for the representation of the number of
`school-seats' that are spatially available to students by mode of
transportation (i.e., motorized and active travel), considering the
school-aged student population by zone. Further, spatial availability
can be represented as school-seats per capita, mathematically equivalent
to the supply-to-demand ratio of the 2SFCA (see appendix in Soukhov et
al. (2023)), which increases the interpretability of the analysis, and
the comparability between years.

@@

Hamilton is a mid-size city (\textasciitilde570,000 pop) in the province
of Ontario, Canada, with rural, suburban and urban characteristics
(Government of Canada 2022a). The city's public English education system
is managed by the Hamilton-Wentworth District School Board (HWDSB) and
the Hamilton-Wentworth Catholic District School Board (HWCDSB). These
two boards, like many in the province have faced fiscal pressures from
the province. For example, in 2013, the HWDSB released its Long Term
Facilities Master Plan (HWDSB 2013), which nominally meant to ensure
``equitable, affordable and sustainable learning facilities''. This plan
indicated that 80\% of its elementary schools (attended by students aged
approximately 5 to 14) would be subject to evaluations, including a
``Pupil Accommodation Review'' (HWDSB 2013). The announcement promptly
led to a public outcry, since the outcome of Pupil Accommodation Reviews
historically translated into school closures and consolidations (Craggs
2013; Seasons 2014b).

School boards in Ontario rely on funding from the province which is
distributed through a formula based on pupil enrollment (Mackenzie 2018;
Irwin and Seasons 2012). However, this funding formula has demonstrably
become insufficient to maintain schools in a state of good repair
(Auditor General of Ontario 2015). Further, existing ``top-up'' programs
to assist in operation and maintenance costs (e.g., School Facilities
Operation and Renewal Grant) were entirely phased out by 2018 (TDSB
2024). Against this backdrop, the Pupil Accommodation Review guidelines
outlined a ``community-focused'' process for assessing the value that a
school provides to students, the community, and the local economy
(Seasons 2014b; Ministry of Education 2006). However in practice,
critical residents felt the outcomes of the Pupil Accommodation Reviews
were predetermined (Thompson, Collins, and Dean 2024): older schools
were slated to be closed, and a wholesome breakdown of costs to repair
and maintain schools was not made available (Kleinhuis 2013). The
process introduced unbalanced incentives for school boards: as a
concerned resident pointed at the time, ``{[}b{]}y closing three
schools, the HWDSB would have a strong case with the Ministry of
Education to receive full funding for a new school'' (Kleinhuis 2013).
The Ministry's framework promotes the closing and consolidation of
schools rather than cost-effective solutions for keeping older schools
open (Kleinhuis 2013).

Hamilton's two school boards (HWDSB and HWCDSB), which together provide
public English education to the majority of school-aged children in
Hamilton (FAO 2023), were particularly impacted by fiscal pressures.
While Accommodation Reviews were a province-wide trend during this
period, Hamilton's school boards underwent an unprecedented number of
reviews compared to boards in other Ontario municipalities. (Seasons
2014a). In the wave of Pupil Accommodation Reviews between 2013 and
2016, 12 of the 147 elementary schools closed in Hamilton. Likely in
part due to the unpopularity of school closures, in 2017, the provincial
government introduced a moratorium, which paused any scheduled Pupil
Accommodation Reviews. This created a backlog of Accommodation Reviews
that HWDSB caught up to in subsequent years, when it closed 12 more
schools between 2017 to 2021 (HWDSB 2023). Though elementary school
locations declined by 10\% between 2011-2021, HWDSB intends to conduct
future Pupil Accommodation Reviews as initially scheduled, waiting until
the end of the moratorium before dates are assigned (HWDSB 2023).

A justification for these Pupil Accommodation Reviews are operational
savings based on ``projected'' reductions in student population and
``under-utilized'' school capacity (Craggs 2012; TDSB 2024; OPSBA 2024).
These policies led to overall reductions in school capacity per student:
school seats in the city declined 5\%--but the impacts were felt
unevenly across the city. It should be noted that although closures
dominated the Pupil Accommodation Review process, some schools were
``consolidated'', that is, a school or two were closed and another
school was expanded or opened to accommodate students from closed
schools. Since school locations decreased proportionally more than
school-seats, elementary schools in Hamilton are on average larger than
they once were.

Existing research indicates that schools have a plethora of benefits not
captured by their maintenance and operational costs (Butler, Kane, and
Cooligan 2019; Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014),
and in this paper we focus on the transportation implications. For
students who live in proximity to schools, closures or consolidations
directly eliminate the potential for active trips, and consequently
physical activity. Closing a school permanently reduces the potential
for active trips and further entrenches reliance on motorized travel.
Additionally, from the perspective of modal equity, school-aged
populations with access to motorized transport gain a speed-time
advantage, providing them with greater flexibility and choice in
accessing school-seats.

The spatial and population-level distribution of these impacts are of
question for this paper. Specifically, the paper's aim is to investigate
how these policies impacted the motorized and non-motorized (active)
school-seat accessibility landscape. We hypothesize that school closure
and consolidation policies reduced the availability of school-seats for
many students. By extension, this loss in accessibility and the
displacement of active trips likely resulted in decreased potential
active travel minutes and increased motorized travel minutes, with these
impacts distributed inequitably across space and populations. To test
these hypotheses, the paper examines the following:

\begin{itemize}
\tightlist
\item
  Did spatial availability change for home-to-school trips following
  school closures and consolidations?
\item
  What is the extent of the loss in walking minutes for these trips?
\item
  How has motorized travel increased as a result?
\item
  Which communities, particularly households with a high prevalence of
  low-income and hence more likely experiencing transport poverty,
  experienced the greatest losses due to this policy?
\end{itemize}

@@

For this study, we use the locations of elementary schools and their
capacities at each of three years: 2011, 2016, and 2021.
